# Analytical Framework

```{r}
#| include: false
source("R/common.R")
```

## Development Potential Scoring

Our methodology evaluates **seven key criteria** across positive attributes and limiting factors to create a composite development score (0-100):

### Six Weighted Components (total: 100%)
```{r scoring-visual, fig.height=4}
#| echo: false

scoring_components <- tibble(
  Component = c("Current Use", "Property Size", "Location Quality", 
                "Financial Need", "Market Potential", "Zoning"),
  Weight = c(25, 20, 20, 15, 10, 10),
  Description = c(
    "Underutilized properties (parking lots, open space)",
    "Optimal size for development (0.5-5 acres scores highest)",
    "Walkability and transit accessibility",
    "Congregation financial health trends",
    "QCT/DDA status and area characteristics",
    "Development-friendly zoning classifications"
  )
) |>
  arrange(desc(Weight))

ggplot(scoring_components, aes(x = reorder(Component, Weight), y = Weight)) +
  geom_col(fill = "#445ca9", alpha = 0.8) +
  geom_text(aes(label = paste0(Weight, "%")), hjust = -0.2, size = 5, 
            color = "#445ca9", fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 30)) +
  labs(
    title = "Development Score Components",
    subtitle = "Weighted contribution to final score (0-100)",
    x = "",
    y = "Weight (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#445ca9"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Plus Constraint Penalties

Environmental and regulatory factors are applied as score deductions:

- Combined flood zone + significant wetlands (>25%): **-75 points**
- Flood zone OR moderate wetlands (>10%): **-40 points**  
- Minor environmental constraints: **-20 points**

*Note: Missing wetland data is treated as neutral (no penalty) to avoid penalizing properties where environmental assessment data is unavailable.*

### Development Classification Systems

Properties are evaluated using two complementary frameworks:

**Goldilocks Size Classification** (for opportunity identification):

- **High Priority**: 1.5-5 acres, minimal constraints, documented value
- **Moderate Priority**: 0.5-1.5 acres, minimal constraints
- **Too Large**: >5 acres (may require phased development or specialized partners)
- **Small Parcel**: <0.5 acres (limited development options)
- **Constrained**: Environmental limitations regardless of size

**Numerical Score Tiers** (for prioritization):

- **Tier 1 (75-100):** High Priority - Immediate development candidates
- **Tier 2 (60-74):** Strong Potential - Near-term opportunities
- **Tier 3 (45-59):** Moderate Potential - Requires creative solutions
- **Tier 4 (30-44):** Limited Potential - Significant barriers exist
- **Tier 5 (<30):** Not Recommended - Unsuitable for development

*The goldilocks classification focuses on the **optimal 0.5-5 acre range** that balances economic viability with manageable complexity. Properties outside this range may have value but require different development strategies.*
