## Developable Land Summary

```{r}
#| include: false
source("R/common.R")
```

### Comprehensive Developability Analysis

```{r developable-summary, echo=FALSE}
#| echo: false

# Summary by congregation with mean development score
developable_by_congregation <- property_profile %>%
  filter(development_score >= 57) %>%
  group_by(congregation_name) %>%
  summarise(
    n_parcels = n(),
    total_acres = sum(area_acres, na.rm = TRUE),
    total_value = sum(assessed_value, na.rm = TRUE),
    avg_attendance = first(avg_attendance),
    avg_members = first(avg_members),
    mean_dev_score = mean(development_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_dev_score))

# Define color based on score
get_score_color <- function(score) {
  case_when(
    score >= 70 ~ "#445ca9",
    score >= 60 ~ "#8baeaa",
    TRUE ~ "#e9ab3f"
  )
}

# Create formatted table
formatted_table <- developable_by_congregation %>%
  mutate(
    total_acres = round(total_acres, 2),
    total_value = scales::dollar(total_value, accuracy = 1),
    avg_attendance = round(avg_attendance, 0),
    avg_members = round(avg_members, 0),
    mean_dev_score = round(mean_dev_score, 1)
  )

# Build kable with row colors
kable(formatted_table,
      col.names = c("Congregation", "Parcels", "Total Acres", "Assessed Value", 
                    "Avg Attendance", "Avg Members", "Mean Dev Score"),
      align = c("l", "r", "r", "r", "r", "r", "r"),
      caption = "Developable Properties by Congregation (Score ≥57)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = TRUE) %>%
  row_spec(which(developable_by_congregation$mean_dev_score >= 70), 
           background = "#445ca9", color = "white") %>%
  row_spec(which(developable_by_congregation$mean_dev_score >= 60 & 
                 developable_by_congregation$mean_dev_score < 70), 
           background = "#8baeaa") %>%
  row_spec(which(developable_by_congregation$mean_dev_score < 60), 
           background = "#e9ab3f")
```

Looking at top developent parcels and their congregations reveals a striking pattern: all 7 congregations with development scores at or above 57/100 report average attendance at or below 20 people, with a median attendance of just 14 people. This universally low attendance suggests these properties represent not just real estate opportunities, but potential candidates for congregational transition—whether through relocation, merger, or strategic closure.

St. Pauls on the Hill Church in Winchester (Frederick County) exemplifies this dynamic. Despite maintaining 60 members on the rolls, the congregation reports zero average attendance. With a development score of 77.6—the highest in the Diocese—and nearly 5 acres of land, St. Pauls represents a flagship opportunity where VEREP engagement could explore whether this congregation is ready for a supported transition that unlocks significant development potential in a priority market.

Similarly, Good Shepherd-of-the-Hills in Albemarle County (63.2) and St. Georges Church in Page County (57.0) both report zero average attendance despite having members on record. These congregations may already be functionally inactive, making them strong candidates for early VEREP outreach to clarify their status and discuss path forward options.

Fairfax County presents a unique opportunity for coordinated engagement. Three congregations—St. Patricks Anglo Vietnamese Church (21 attendance), Holy Cross Korean Episcopal Church (25 attendance), and All Saints Sharon Chapel (28 attendance)—are all operating with modest but active congregations. Given their geographic proximity and similar scale, VEREP could explore whether a merger strategy might strengthen congregational life while freeing valuable property for development.

Grace Church Emmanuel Parish in Fauquier County holds the largest parcels among top developable congregations but averages only 14 in attendance with 38 members. This table only reflects the two highest scored of their parcels, or 6.4 acres out of their 14.7 total acres across 3 parcels. As a multi-parcel congregation, Grace Church may be able to consolidating its footprint, potentially releasing surplus land for development while maintaining a right-sized worship presence.

::: {.callout-caution}
Congregation data also has its flaws.

Aquia Church in Stafford County, for example, claims 1,136 members but zero average attendance across 5 parcels. This discrepancy warrants VEREP engagement to understand the congregation's actual status and whether the membership figure reflects historical rolls rather than current participation. 
:::



