## Development Opportunity Landscape

```{r}
#| include: false
source("R/common.R")
```

Of the 104 parcels pulled from the VEREP GIS tool for examination, 4 were immediately removed from development opportunity consideration due to their size. These parcels were all under 0.1 acres -- likely cemetery plots, parking strips, or slivers of land.  Three of these tiny parcels belong to the same congregation, so while not candidates for independent development, they may hold value as part of a larger assemblage.

```{r}
#| echo: false
#| results: asis

library(kableExtra)

small_parcels <- property_profile %>%
  filter(area_acres <= 0.1) %>%
  select(name, congregation_name, scounty, area_acres) %>%
  mutate(
    sq_feet = round(area_acres * 43560, 0),
    sq_feet = format(sq_feet, big.mark = ","),
    area_acres = sprintf("%.03f", area_acres)
  )

small_parcels %>%
  kable(
    col.names = c("Property", "Congregation", "County", "Acres", "Sq Feet"),
    align = c("l", "l", "l", "r", "r")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  print()


```

This means ultimately 100 parcels were scored and ranked for development opportuntity: 

```{r tier-distribution-visual}
#| echo: false

tier_data <- scored_parcels |>
  count(development_potential) |>
  mutate(
    pct = n / sum(n) * 100,
    color = case_when(
      development_potential == "High" ~ "#445ca9",
      development_potential == "Moderate" ~ "#8baeaa",
      development_potential == "Constrained" ~ "#e9ab3f",
      development_potential == "Too Large" ~ "#e76f52",
      development_potential == "Small Parcel" ~ "#909090",
      TRUE ~ "#cccccc"
    )
  )

ggplot(tier_data, aes(x = reorder(development_potential, n), y = n, fill = color)) +
  geom_col() +
  geom_text(aes(label = paste0(n, "\n(", round(pct, 1), "%)")), 
            vjust = 1.5, color = "white", size = 4, fontface = "bold") +
  scale_fill_identity() +
  labs(
    title = "Distribution of Properties by Development Tier",
    subtitle = paste0("Total Properties Analyzed: ", sum(tier_data$n)),
    x = "",
    y = "Number of Properties"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#445ca9"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    panel.grid.major.x = element_blank()
  )
```
 
Nearly half the portfolio (47%) falls within the goldilocks range of 0.5-5 acres with minimal constraints, representing actionable development opportunities. The remaining 53% face size limitations—either too small for economical development (39%) or too large for congregation-scale management (11%). Environmental constraints affect only 3% of properties, indicating the portfolio is largely unconstrained by wetlands or flood risk. However, this could be due to our choice to process missing wetland data as neutral, and should be carefully examined when beginning due diligence for the high-potential sites. 