## Development Opportunity Landscape

```{r}
#| include: false
source("R/common.R")
```

Of the 104 parcels pulled from the VEREP GIS tool for examination, 4 were immediately removed from development opportunity consideration due to their size. These parcels were all under 0.1 acres -- likely cemetery plots, parking strips, or slivers of land.  Three of these tiny parcels belong to the same congregation, so while not candidates for independent development, they may hold value as part of a larger assemblage.

```{r}
#| echo: false
#| results: asis

library(kableExtra)

small_parcels <- property_profile %>%
  filter(area_acres <= 0.1) %>%
  select(name, congregation_name, scounty, area_acres) %>%
  mutate(
    sq_feet = round(area_acres * 43560, 0),
    sq_feet = format(sq_feet, big.mark = ","),
    area_acres = sprintf("%.03f", area_acres)
  )

small_parcels %>%
  kable(
    col.names = c("Property", "Congregation", "County", "Acres", "Sq Feet"),
    align = c("l", "l", "l", "r", "r")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  print()

```

**This means ultimately only 100 parcels were scored and ranked for development opportuntity.**

```{r tier-distribution-visual}
#| echo: false

tier_data <- scored_parcels |>
  count(development_potential) |>
  mutate(
    pct = n / sum(n) * 100,
    color = case_when(
      development_potential == "High Potential" ~ "#445ca9",
      development_potential == "Moderate Potential" ~ "#8baeaa",
      development_potential == "Low Potential - Constraints" ~ "#e76f52",
      development_potential == "Low Potential - Small Parcel" ~ "#e9ab3f",
      TRUE ~ "#cccccc"
    )
  )

ggplot(tier_data, aes(x = reorder(development_potential, n), y = n, fill = color)) +
  geom_col() +
  geom_text(aes(label = paste0(n, "\n(", round(pct, 1), "%)")), 
            vjust = 1.5, color = "white", size = 4, fontface = "bold") +
  scale_fill_identity() +
  labs(
    title = "Distribution of Properties by Development Tier",
    subtitle = paste0("Total Properties Analyzed: ", sum(tier_data$n)),
    x = "",
    y = "Number of Properties"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#445ca9"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    panel.grid.major.x = element_blank()
  )
```

Of the 100 properties analyzed with congregations under 30 Sunday attendees, **46 properties (47%) fall within the ideal development range** of chosen criteria.

**High Potential**: 23 properties (23.5%) — 1.5+ acres with optimal development characteristics
**Moderate Potential**: 23 properties (23.5%) — 0.5-1.5 acres with development potential

The remaining 53% face size or constraint limitations: **35 properties (35.7%) are small parcels** under 0.5 acres offering limited standalone development options, and **17 properties (17.3%) are environmentally constrained** by wetlands, flood zones, or high natural hazard risk.

This distribution reveals that nearly half the portfolio presents actionable development opportunities—a strong finding for an institutional portfolio. The primary limitation is parcel size: over a third of properties are simply too small for economical standalone development, though they may still support ADUs, small-scale infill, or land assembly strategies.