# Executive Summary {.unnumbered}

This analysis examines properties in the Episcopal Diocese of Virginia with fewer than 30 Sunday attendees, focusing on development potential through five key metrics: land value, zoning/density, acreage, environmental limitations, and Qualified Census Tract (QCT) status. This report was done for VEREP (Virginia Episcopal Real Estate Partners) in conjunction with case study projects HDAdvisors is facilitating.

## At a Glance: Top 10 Properties

```{r}
#| echo: false
source("R/common.R")
source("R/app.R")

```

```{r}
#| label: tbl-top-10
#| tbl-cap: "Top 10 Development Opportunities"
#| echo: false

top_10 |>
  select(rank, site_address, site_city, area_acres, development_score, development_tier) |>
  kable(
    col.names = c("Rank", "Address", "City", "Acres", "Score", "Tier"),
    digits = c(0, 0, 0, 2, 1, 0),
    align = c("c", "l", "l", "r", "r", "l")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  column_spec(1, bold = TRUE, width = "3em") |>
  column_spec(5, color = "white", 
              background = spec_color(top_10$development_score, 
                                      scale_from = c(0, 100),
                                      palette = "YlOrRd",
                                      direction = -1))
```

```{r}
#| label: fig-top-10-scores
#| fig-cap: "Development Scores for All Properties"
#| fig-height: 5



scored_parcels |>
  filter(!is.na(development_score)) |>
  ggplot(aes(x = development_score, fill = development_potential)) +
  geom_histogram(binwidth = 5, color = "white") +
  scale_fill_manual(values = c(
    "High Potential" = "#445ca9",
    "Moderate Potential" = "#8baeaa",
    "Low Potential - Constraints" = "#e9ab3f",
    "Insufficient Data" = "#e76f52"
  )) +
  labs(
    title = "Distribution of Development Scores",
    x = "Development Score",
    y = "Number of Properties",
    fill = "Tier"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#445ca9"),
    legend.position = "bottom"
  )
```

## Key Findings

### Portfolio Summary

```{r}
#| label: portfolio-summary
#| echo: false
#| results: asis

# Calculate summary statistics
total_props <- nrow(scored_parcels)
high_potential <- sum(scored_parcels$development_potential == "High Potential", na.rm = TRUE)
moderate_potential <- sum(scored_parcels$development_potential == "Moderate Potential", na.rm = TRUE)
strong_potential <- high_potential + moderate_potential
total_acres <- sum(scored_parcels$area_acres, na.rm = TRUE)
top_10_acres <- sum(top_10$area_acres, na.rm = TRUE)

cat("::: {layout-ncol=4}\n\n")

cat("::: {.callout-note appearance='minimal'}\n")
cat("## ", total_props, "\n", sep = "")
cat("Properties Analyzed\n")
cat(":::\n\n")

cat("::: {.callout-tip appearance='minimal'}\n")
cat("## ", strong_potential, "\n", sep = "")
cat("Strong Potential (Tier 1-2)\n")
cat(":::\n\n")

cat("::: {.callout-warning appearance='minimal'}\n")
cat("## ", round(top_10_acres, 1), "\n", sep = "")
cat("Top 10 Developable Acres\n")
cat(":::\n\n")

cat("::: {.callout-important appearance='minimal'}\n")
cat("## ", round(total_acres, 0), "\n", sep = "")
cat("Total Portfolio Acres\n")
cat(":::\n\n")

cat(":::\n")
```

### Significant Underutilized Assets

```{r}
#| echo: false
#| results: asis

parking_count <- sum(scored_parcels$use_parking, na.rm = TRUE)
open_space_count <- sum(scored_parcels$use_open_space, na.rm = TRUE)

cat("Our analysis identified **", strong_potential, " properties** with substantial development potential.\n\n", sep = "")
```

**Primary opportunities include:**

-   **Parking lots** (highest development scores): `r sum(scored_parcels$use_parking, na.rm = TRUE)` properties with underutilized parking suitable for structured parking with mixed-use above
-   **Open space parcels**: `r sum(scored_parcels$use_open_space, na.rm = TRUE)` well-located vacant land parcels ideal for ground lease arrangements
-   **Properties with declining congregations**: Sites where development can provide financial stability

### Geographic Distribution

```{r}
#| label: fig-geographic-dist
#| fig-cap: "Properties by County"
#| fig-height: 4

county_summary <- scored_parcels |>
  count(site_county, name = "n_properties") |>
  arrange(desc(n_properties)) |>
  slice_head(n = 10)

ggplot(county_summary, aes(x = reorder(site_county, n_properties), y = n_properties)) +
  geom_col(fill = "#445ca9", alpha = 0.8) +
  geom_text(aes(label = n_properties), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Top 10 Counties by Property Count",
    x = "",
    y = "Number of Properties"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#445ca9"),
    panel.grid.major.y = element_blank()
  )
```

Development opportunities are distributed across Virginia, with concentrations in:

-   Urban/suburban areas with strong walkability scores
-   Markets with favorable demographics and LIHTC eligibility
-   Transit-accessible locations supporting reduced parking requirements

### Financial Sustainability Imperative

::: callout-warning
## Critical Finding

```{r}
#| echo: false
#| results: asis

declining_count <- sum(scored_parcels$financial_need >= 2, na.rm = TRUE)
cat("**", declining_count, " properties** are associated with congregations showing declining financial trends. Development of identified properties is essential for long-term financial sustainability and mission fulfillment.\n", sep = "")
```
:::

## Development Tier Overview

```{r}
#| label: fig-tier-distribution
#| fig-cap: "Distribution of Properties by Development Potential"
#| fig-height: 4

tier_summary <- scored_parcels |>
  count(development_potential) |>
  mutate(
    pct = n / sum(n) * 100,
    tier_order = case_when(
      development_potential == "High Potential" ~ 1,
      development_potential == "Moderate Potential" ~ 2,
      development_potential == "Low Potential - Constraints" ~ 3,
      development_potential == "Insufficient Data" ~ 4,
      TRUE ~ 5
    )
  ) |>
  arrange(tier_order)

ggplot(tier_summary, aes(x = reorder(development_potential, -tier_order), y = n,
                          fill = development_potential)) +
  geom_col() +
  geom_text(aes(label = paste0(n, " (", round(pct, 1), "%)")), 
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c(
    "High Potential" = "#445ca9",
    "Moderate Potential" = "#8baeaa",
    "Low Potential - Constraints" = "#e9ab3f",
    "Insufficient Data" = "#e76f52"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "",
    y = "Number of Properties"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#445ca9"),
    axis.text.x = element_text(angle = 25, hjust = 1),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  )
```

## Next Steps

1.  **Priority Assessment**: Review top-ranked properties for immediate action
2.  **Feasibility Studies**: Conduct detailed feasibility analysis for top candidates
3.  **Stakeholder Engagement**: Engage relevant stakeholders for high-priority sites
4.  **Financial Planning**: Develop funding strategies and timelines
