---
title: "VEREP Property Development Analysis"
author: "HDAdvisors"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
  pdf:
    toc: true
    toc-depth: 3
    geometry:
      - margin=1in
    mainfont: "Lato"
    sansfont: "Lato"
execute:
  warning: false
  message: false
---

```{css}
#| echo: false

@import url('https://fonts.googleapis.com/css2?family=Roboto+Slab:wght@400;700&family=Lato:wght@300;400;700&display=swap');

/* Apply HDA fonts */
body {
  font-family: 'Lato', sans-serif;
  color: #333333;
}

h1, h2, h3, h4, h5, h6 {
  font-family: 'Roboto Slab', serif;
  color: #445ca9;
}

/* Style callouts with HDA colors */
.callout-note {
  border-left-color: #445ca9 !important;
}

.callout-warning {
  border-left-color: #e76f52 !important;
}

.callout-tip {
  border-left-color: #8baeaa !important;
}

/* Style links */
a {
  color: #445ca9;
}

a:hover {
  color: #e9ab3f;
}
```

## Executive Summary
- Brief overview of the top 10 developable properties
- Key findings and recommendations
- Quick reference table/map

## 1. Background & Data Sources
### 1.1 VEREP Parcel Dataset Methodology
[Condensed version of the CGS methodology summary]

### 1.2 Congregation Statistics
- What metrics are available (attendance, pledge, membership)
- Time period covered (2014-2023)
- Data quality considerations

## 2. Analytical Approach

### 2.1 Solving the One-to-Many Join Problem
```{r congregation-join}
# Your approach to handling parcels with multiple congregations
# Document decisions: average metrics? primary congregation? all combinations?
```

### 2.2 Defining "Developability"
Create a scoring framework based on:
```{r developability-criteria}
# Property characteristics:
# - Parcel size (adequate for development?)
# - Current use (underutilized parking lots? vacant land?)
# - Zoning favorability
# - Not in flood zones/wetlands

# Financial indicators:
# - Assessed value vs. potential
# - Congregation financial health (declining pledges = need for income?)

# Location factors:
# - Walkability score
# - Transit access
# - Median income (market demand)
# - Urban vs. rural

# Regulatory ease:
# - Not in historic districts
# - Not encumbered by easements
# - LIHTC qualified census tract (for affordable housing incentives)
```

### 2.3 Scoring Methodology
```{r scoring-system}
# Weighted composite score
# Normalize each criterion to 0-100 scale
# Apply weights based on VEREP priorities
# Calculate total development potential score
```

## 3. Analysis & Results

### 3.1 Congregation-Parcel Integration
```{r join-results}
# Show how you resolved the many-to-many issue
# Summary statistics of joined dataset
# Data quality checks

```

### 3.2 Development Potential Scoring
```{r score-distribution}
# Distribution of scores across all parcels
# Visualization of scoring components
# Identify high-potential clusters
```

### 3.3 Top 10 Developable Properties
```{r top-10-analysis}
# Detailed profiles of top 10
# Maps showing locations
# Comparative analysis
```

## 4. Deep Dive: Top Properties

### Property 1: [Name/Address]
- Parcel characteristics
- Congregation context (health metrics, trends)
- Development opportunities
- Considerations/constraints

[Repeat for each top 10]

## 5. Recommendations & Next Steps

## Appendices
- Full methodology
- Data dictionary
- Additional maps/tables