## Geographic Distribution
```{r}
#| include: false
source("R/common.R")
source("R/app.R")
```

## Interactive Map

::: {.content-visible when-format="html"}
```{r all-properties-map-html}
#| echo: false
#| eval: !expr knitr::is_html_output()

pal <- colorNumeric(
  palette = "YlOrRd",
  domain = scored_parcels$development_score
)

leaflet(scored_parcels) |>

addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    radius = ~sqrt(area_acres) * 3,
    fillColor = ~pal(development_score),
    fillOpacity = 0.7,
    color = "white",
    weight = 1,
    popup = ~paste0(
      "<div style='font-family: Lato, sans-serif; min-width: 250px;'>",
      "<p style='margin: 5px 0;'><b>", site_address, "</b><br>",
      site_city, ", ", site_county, " County</p>",
      "<hr style='margin: 10px 0;'>",
      "<p style='margin: 5px 0;'><b>Development Score:</b> ", round(development_score, 1), "/100<br>",
      "<b>Size:</b> ", round(area_acres, 2), " acres<br>",
      "<b>Tier:</b> ", development_tier, "</p>",
      if_else(!is.na(name), 
              paste0("<hr style='margin: 10px 0;'><p style='margin: 5px 0;'><b>Congregation:</b> ", name, "</p>"), 
              ""),
      "</div>"
    ),
    label = ~paste0(site_address, " (", round(development_score, 1), ")")
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~development_score,
    title = "Development<br>Score",
    opacity = 1
  ) |>
  setView(lng = mean(scored_parcels$lon, na.rm = TRUE), lat = mean(scored_parcels$lat, na.rm = TRUE), zoom = 7)
```
:::

Development opportunities are distributed broadly across the Diocese, with the color gradient revealing where the highest-scoring properties cluster. **High-potential properties (scores 60–80)**, shown in deep red, are concentrated in the **Northern Virginia/DC corridor** near Alexandria and Reston, as well as scattered throughout the **Shenandoah Valley** from Winchester through Harrisonburg. These sites score well due to favorable zoning, strong market conditions, and fewer environmental constraints.

The **Fredericksburg region** stands out for its large-acreage properties—represented by the largest circles on the map—though these score in the **moderate range (40–50)**. Their size represents significant opportunity, but factors such as zoning or environmental considerations may require additional due diligence before development.

Other potential development opportunities are spread across the Diocese, including areas near Charlottesville and Richmond. These sites may benefit from rezoning efforts or phased development strategies.

This geographic diversity is an asset: the Diocese can pursue high-score "quick wins" in Northern Virginia while simultaneously advancing larger but more complex opportunities in growth corridors like Fredericksburg. The portfolio supports both near-term revenue generation and long-term strategic development.

::: {.content-visible when-format="pdf"}

*The interactive map is available in the HTML version of this report.*

:::