# Property Profiles

```{r}
#| include: false
source("R/common.R")
source("R/app.R")
```


## Overview

In this analysis, the top 10 properties all scored the same, which means they are listed in no particular order, but as the top group of high-possibility sites.

The following section provides detailed profiles for each of the top 10 development opportunities. Each profile includes:

-   Site characteristics and current use
-   Congregation context (where applicable)
-   Development opportunities and strategies
-   Key considerations and next steps

```{r property-profiles, results='asis'}
#| echo: false 

create_property_profile <- function(rank_num) {
  property <- top_10 |> filter(rank == rank_num)
  
  if(nrow(property) == 0) return(NULL)
  
  cat("\n\n## Property #", rank_num, " {.unnumbered}\n\n", sep = "")
  cat("### ", property$site_address, "\n", sep = "")
  cat("**", property$site_city, ", ", property$site_county, " County**\n\n", sep = "")
  
  tier_type <- case_when(
    property$development_score >= 75 ~ "important",
    property$development_score >= 60 ~ "tip",
    TRUE ~ "note"
  )
  
  cat(":::{.callout-", tier_type, "}\n", sep = "")
  cat("## Development Score: **", round(property$development_score, 1), "/100**\n\n", sep = "")
  cat("**", property$development_tier, "**\n", sep = "")
  cat(":::\n\n")
  
  cat("::: {layout-ncol=3}\n")
  cat("::: {}\n**", round(property$area_acres, 2), " acres**  \n", sep = "")
  cat("Property Size\n:::\n")
  
  cat("::: {}\n**", round(property$walkability_score, 1), "/100**  \n", sep = "")
  cat("Walkability\n:::\n")
  
  cat("::: {}\n**", 
      ifelse(!is.na(property$avg_attendance), 
             paste0(round(property$avg_attendance), " people"), "N/A"), "**  \n", sep = "")
  cat("Avg Attendance\n:::\n")
  cat(":::\n\n")
  
  cat("#### Site Characteristics\n\n")
  cat("- **Current Uses:**")
  uses <- c()
  if(!is.na(property$use_church) && property$use_church) uses <- c(uses, "Church building")
  if(!is.na(property$use_parking) && property$use_parking) uses <- c(uses, "Parking lot")
  if(!is.na(property$use_open_space) && property$use_open_space) uses <- c(uses, "Open space")
  if(length(uses) > 0) {
    cat(" ", paste(uses, collapse = ", "), "\n", sep = "")
  } else {
    cat(" Mixed use\n")
  }
  
  if(!is.na(property$name)) {
    cat("\n#### Congregation: ", property$name, "\n\n", sep = "")
    
    if(!is.na(property$pct_change_pledge)) {
      change_dir <- ifelse(property$pct_change_pledge < 0, "â†“ decreased", "â†‘ increased")
      cat("**Financial Trend:** Plate & pledge has ", change_dir, " by **", 
          abs(round(property$pct_change_pledge, 1)), "%** over the past decade.\n\n", sep = "")
    }
  }
  
  cat("#### Development Opportunities\n\n")
  if(!is.na(property$use_parking) && property$use_parking) {
    cat("ðŸ…¿ï¸ **Prime opportunity:** Underutilized parking can support structured parking with mixed-use above\n\n")
  }
  if(!is.na(property$area_acres) && property$area_acres > 1) {
    cat("ðŸ“ **Substantial site:** Large parcel enables phased development or creative master planning\n\n")
  }
  if(!is.na(property$walkability_score) && property$walkability_score > 10) {
    cat("ðŸš¶ **Walkable location:** Transit access supports reduced parking requirements and higher density\n\n")
  }
  
  cat("#### Recommended Next Steps\n\n")
  cat("1. Site feasibility study and geotechnical analysis\n")
  cat("2. Congregation leadership engagement\n")
  cat("3. Zoning review and density bonus evaluation\n")
  if(!is.na(property$financial_need) && property$financial_need >= 2) {
    cat("4. Ground lease financial modeling\n")
  }
  
  cat("\n---\n\n")
}

# Generate all profiles
for(i in 1:10) {
  create_property_profile(i)
}
```


## Interactive ADU Placement Tool

For an interactive experience, [**launch the ADU Placement Dashboard**] 
<iframe 
  src="https://mted.shinyapps.io/shiny-app/?tab=adu"
  width="100%" 
  height="600px"
  frameborder="0">
</iframe>
