---
title: "VEREP Development Opportunities"
subtitle: "Executive Briefing"
author: "HDAdvisors"
date: today
format:
  pptx:
    reference-doc: custom-template.pptx
  revealjs:  # alternative: HTML slides
    theme: serif
    transition: slide
---

# Strategic Overview

## Portfolio At-A-Glance

- **357 properties** analyzed across diocese
- **87 properties** with strong development potential
- **31.4 acres** in top 10 opportunities
- **6 sites** supporting financially stressed congregations

---

# Key Findings

## Significant Underutilized Assets

::: {.columns}
::: {.column width="50%"}
**Primary Opportunities:**

- Parking lots
- Open space parcels
- Declining congregations
:::

::: {.column width="50%"}
**Tier 1-2 Properties:**

- 87 parcels
- Strong scores (60-100)
- Immediate potential
:::
:::

---

# Development Priorities

## Top 5 Sites
```{r}
#| echo: false
top_5_table <- top_10 %>%
  slice_head(n = 5) %>%
  select(rank, site_address, site_city, area_acres, development_score)

kable(top_5_table, 
      col.names = c("Rank", "Address", "City", "Acres", "Score"))
```

---

# Scoring Framework

## Six Key Dimensions
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5

# Your scoring visual from the main document
scoring_components <- tibble(
  Component = c("Property Size", "Current Use", "Location Quality", 
                "Financial Need", "Market Potential", "Zoning"),
  Weight = c(20, 25, 20, 15, 10, 10)
)

ggplot(scoring_components, aes(x = reorder(Component, Weight), y = Weight)) +
  geom_col(fill = "#445ca9", alpha = 0.8) +
  coord_flip() +
  labs(title = "Component Weights", y = "Weight (%)") +
  theme_minimal(base_size = 14)
```

---

# Geographic Distribution

## Interactive Map
```{r}
#| echo: false
# Your leaflet map code
```

---

# Conclusion

## Key Takeaways

::: {.incremental}
- Substantial untapped development value in portfolio
- 87 properties ready for strategic partnerships
- Top 10 sites = 31+ developable acres
- Immediate action: feasibility studies on top 3
:::

## Questions?